The Talking Heads Experiment

### 2015-04-27 chap1/chap1.tex
hyphenation, punctuation, positioning, etc for chapters 1-6
fixed remaining typos

A fourth radical idea concerns the nature and origins of 
grammar. Rather than invoking the need of a highly 
specialised genetically determined language 
organ,\footnote{
As strongly argued by Chomsky in various writings. 
See for example \cite{Chomsky:1968}. 
Praeger, New York. Even though Chomsky argues for 
an innate language acquisition device he has expressed
scepticism about evolution by natural selection
as an explanation for the 
origin of this device (and therefore of language). 
A genetic theory of language evolution 
has been suggested by \cite{Pinker:1994}.} 
I believe that grammar spontaneously arises when 
generic capabilities to categorise reality, store past
events in terms of abstract schemas, remember associations
between events, etc., reach a critical level and are
applied to language itself. These capabilities are 
relevant across many different cognitive domains. 
In order to store linguistic
experiences, human memories spontaneously structure them, thus 
introducing abstract schemas, internal categories, and 
roles that substructures can play in schemas.
These organisational elements then become externalised. 
Categories are marked by enriching
the form of words, schema boundaries are marked by 
imposing patterns on the expression of a schema, 
roles are marked by assigning
them to specific positions in a 
pattern. This externalisation 
increases the reliability in communication because
it reduces ambiguities and supplies additional context. It 
also aids the stable transmission of the language from one
generation to the next because the language learner
gets additional hints to guess the meaning and function of unknown
words and constructions. 
Once such structuring devices are in place, they help 
to increase the expressive power of the language. The 
complexity of possible language interactions can increase, 
and words or word groups which had multiple usages
can become specialised for particular semantic functions.

@@

\begin{figure}[htbp]
  \centerline{\includegraphics[width=.60\textwidth]{chap1/figs/robot.pdf}}
\caption{Example of a {\it Lego vehicle} built by Tim Smithers based on Lego bricks, a sensori-motor processing board, 
and a variety of 
sensors and actuators. We developed these robots in the early nineties for exploring a behaviour-oriented 
approach to robotics.}
\label{f:plate3}
\end{figure}

@@

To investigate the role of the environment in shaping the
evolving sensori-motor capacities of these robots, 
we built various robotic 
ecosystems in which robots could recharge themselves but also
had to work for their living by 
dimming lights that took away energy from the total energy 
flowing in their ecosystem (see \figref{f:plate6}). Visitors
to our lab could see robots helping each other or 
engaging in fierce competition for
the resources available for survival. In all of this
research, we tried to see how far behaviours would
autonomously evolve, in other words we tried to find 
mechanisms by which the robots would bootstrap themselves 
towards greater sensori-motor complexity. One of the 
main lessons from these experiments was that 
explanations for cognition lie partly outside the brain 
of the individual agent: The environment, the body, the 
sensori-motor apparatus and the behaviour of the other 
agents all partly shaped their capacities and 
further development.\footnote{This approach is also known as {\it situated cognition}. 
\cite{Clancey:1997}, see also: \cite{Varela:1991}.}

@@

In the summer of 1995, a clear breakthrough occurred.  
I was working as a visiting researcher in the Sony
Computer Science Laboratory in Tokyo, invited by its director
Mario Tokoro. Reflecting on our 
experiments from a distance, two new ideas occurred to me. 
First of all, language may have been the missing key
in the initial experiments. Language may 
be a necessary route by which the human cognitive system
bootstraps itself autonomously, in tight interaction with the
environment and aided by a community of other
language speakers. This suggested that if we wanted
to have emergent
forms of cognitive intelligence, we needed to go the same 
route. Second, the principles and mechanisms that had been
pouring out of the study of complexity had to be
relevant to understanding the origins and evolution of language,
because they provided generic explanations for how complexity 
may emerge. These principles include 
self-organisation, structural coupling, 
selectionism, level formation, 
and many others (\citealt{Nicolis:1989}). The field of {\it artificial life} brings together researchers
exploring the insights of complex systems with computational
and robotic experiments (see: \citealt{Langton:1995}).

Together with 
other researchers interested in the then-arising field of `artificial life', I had 
already been simulating path formation in 
ant societies and other biological phenomena exhibiting 
an emergence of complexity. It dawned on me that the
importance of these mechanisms for bootstrapping intelligence and
language might be much greater than 
thought so far.\footnote{The following reference provides a general survey of similar work in 
the area of lexicon formation: \cite{Steels:97b}. 
A representative sample of work on syntax, is \cite{Briscoe:1999}.}

\clearpage
Back in Brussels, an extremely exciting but very 
intense period of research started as I tried to 
apply these principles to language
and subject them to experimental 
scrutiny.\footnote{The earliest papers on these mechanisms are in: \cite{Steels:95b} and 
\cite{Steels:96a}.}
We very quickly built a first 
prototype of a {\it Talking Head} with our own hardware
and low-level software (see \figref{f:plate7}) 
and experimented with language games on mobile 
robotic agents.\footnote{The electronics and tracking software for this 
active camera were built by Tony Belpaeme. The mobile
robot experiments were conducted with Paul Vogt. 
Other early work on the grounding and autonomous acquisition of 
language-like communication systems by robotic systems is 
described in \cite{Steels:97g}. See also: \cite{Billard:1998}.}
New fundamental insights and discoveries emerged almost daily. 
The first experiments, particularly in grounding
language on real robots, were extremely difficult but we 
were clearly making steady progress.

\begin{figure}[htbp]
  \centerline{\includegraphics[width=.70\textwidth]{chap1/figs/eye.pdf}}
\caption{First prototype of a {\it Talking Head} camera with 
associated electronics built by Tony Belpaeme. The camera is capable to track moving images. }
\label{f:plate7}
\end{figure}

@@

The goal of the Talking Heads experiment is not to demonstrate 
an artificial intelligence with the same capacities as 
human intelligence, but to perform
scientific experiments so as to examinine aspects 
of a theory of the origins of language and meaning. 
However, as in Turing's proposal, the public should be 
the ultimate judge whether cognitive phenomena are taking
place, and the artificial agents should be able to play their role
in interaction with humans. 
Sound scientific methodology requires that 
the experimental apparatus is a white box which can be fully probed 
by anyone who wants, that the experiments are repeatable, 
and that the phenomena that are generated 
(for example, the complexity of the lexicons or 
grammars) can be 
compared by anyone to human cognitive phenomena in order to gage 
their similarity and thus their relevance for understanding
human cognition.

@@

The {\it Laboratory for cognitive robots and 
teleportation} that housed the 
Talking Heads experiment also contained  
a documentation room in which the audience could 
get additional background information and 
provide feedback and commentary 
on the experiment. We created a website that was accessible 
worldwide through the Internet. This allowed viewers from anywhere
in the world to follow the dialogs, inspect the lexicons
and ontologies (sets of perceptually grounded concepts) 
of the robots, and even interact remotely with the 
physical robots, playing language games 
and doing their own experiments. We also added
other physical sites in Tokyo, Brussels, Paris, Amsterdam, 
San Jose (US) and other places, to increase the environmental complexity that 
the agents could experience.

>>>>

A fourth radical idea concerns the nature and origins of 
grammar. Rather than invoking the need of a highly 
specialised genetically determined language 
organ,\footnote{
As strongly argued by Chomsky in various writings. 
See for example \cite{Chomsky:1968}. Even though Chomsky argues for 
an innate language acquisition device he has expressed
scepticism about evolution by natural selection
as an explanation for the 
origin of this device (and therefore of language). 
A genetic theory of language evolution 
has been suggested by \cite{Pinker:1994}.} 
I believe that grammar spontaneously arises when 
generic capabilities to categorise reality, store past
events in terms of abstract schemas, remember associations
between events, etc., reach a critical level and are
applied to language itself. These capabilities are 
relevant across many different cognitive domains. 
In order to store linguistic
experiences, human memories spontaneously structure them, thus 
introducing abstract schemas, internal categories, and 
roles that substructures can play in schemas.
These organisational elements then become externalised. 
Categories are marked by enriching
the form of words, schema boundaries are marked by 
imposing patterns on the expression of a schema, 
roles are marked by assigning
them to specific positions in a 
pattern. This externalisation 
increases the reliability in communication because
it reduces ambiguities and supplies additional context. It 
also aids the stable transmission of the language from one
generation to the next because the language learner
gets additional hints to guess the meaning and function of unknown
words and constructions. 
Once such structuring devices are in place, they help 
to increase the expressive power of the language. The 
complexity of possible language interactions can increase, 
and words or word groups which had multiple usages
can become specialised for particular semantic functions.

@@

\begin{figure}[htbp]
  \centerline{\includegraphics[width=.60\textwidth]{chap1/figs/robot.pdf}}
\caption{Example of a {\itshape Lego vehicle} built by Tim Smithers based on Lego bricks, a sensori-motor processing board, 
and a variety of 
sensors and actuators. We developed these robots in the early nineties for exploring a behaviour-oriented 
approach to robotics.}
\label{f:plate3}
\end{figure}

@@

To investigate the role of the environment in shaping the
evolving sensori-motor capacities of these robots, 
we built various robotic 
ecosystems in which robots could recharge themselves but also
had to work for their living by 
dimming lights that took away energy from the total energy 
flowing in their ecosystem (see \figref{f:plate6}). Visitors
to our lab could see robots helping each other or 
engaging in fierce competition for
the resources available for survival. In all of this
research, we tried to see how far behaviours would
autonomously evolve, in other words we tried to find 
mechanisms by which the robots would bootstrap themselves 
towards greater sensori-motor complexity. One of the 
main lessons from these experiments was that 
explanations for cognition lie partly outside the brain 
of the individual agent: The environment, the body, the 
sensori-motor apparatus and the behaviour of the other 
agents all partly shaped their capacities and 
further development.\footnote{This approach is also known as {\itshape situated cognition}. 
\cite{Clancey:1997}, see also: \cite{Varela:1991}.}

@@

In the summer of 1995, a clear breakthrough occurred.  
I was working as a visiting researcher in the Sony
Computer Science Laboratory in Tokyo, invited by its director
Mario Tokoro. Reflecting on our 
experiments from a distance, two new ideas occurred to me. 
First of all, language may have been the missing key
in the initial experiments. Language may 
be a necessary route by which the human cognitive system
bootstraps itself autonomously, in tight interaction with the
environment and aided by a community of other
language speakers. This suggested that if we wanted
to have emergent
forms of cognitive intelligence, we needed to go the same 
route. Second, the principles and mechanisms that had been
pouring out of the study of complexity had to be
relevant to understanding the origins and evolution of language,
because they provided generic explanations for how complexity 
may emerge. These principles include 
self-organisation, structural coupling, 
selectionism, level formation, 
and many others (\citealt{Nicolis:1989}). The field of {\itshape artificial life} brings together researchers
exploring the insights of complex systems with computational
and robotic experiments (see: \citealt{Langton:1995}).

Together with 
other researchers interested in the then-arising field of `artificial life', I had 
already been simulating path formation in 
ant societies and other biological phenomena exhibiting 
an emergence of complexity. It dawned on me that the
importance of these mechanisms for bootstrapping intelligence and
language might be much greater than 
thought so far.\footnote{The following reference provides a general survey of similar work in 
the area of lexicon formation: \cite{Steels:97b}. 
A representative sample of work on syntax is \cite{Briscoe:1999}.}

\clearpage
Back in Brussels, an extremely exciting but very 
intense period of research started as I tried to 
apply these principles to language
and subject them to experimental 
scrutiny.\footnote{The earliest papers on these mechanisms are in: \cite{Steels:95b} and 
\cite{Steels:96a}.}
We very quickly built a first 
prototype of a {\itshape Talking Head} with our own hardware
and low-level software (see \figref{f:plate7}) 
and experimented with language games on mobile 
robotic agents.\footnote{The electronics and tracking software for this 
active camera were built by Tony Belpaeme. The mobile
robot experiments were conducted with Paul Vogt. 
Other early work on the grounding and autonomous acquisition of 
language-like communication systems by robotic systems is 
described in \cite{Steels:97g}. See also: \cite{Billard:1998}.}
New fundamental insights and discoveries emerged almost daily. 
The first experiments, particularly in grounding
language on real robots, were extremely difficult but we 
were clearly making steady progress.

\begin{figure}[htbp]
  \centerline{\includegraphics[width=.70\textwidth]{chap1/figs/eye.pdf}}
\caption{First prototype of a {\itshape Talking Head} camera with 
associated electronics built by Tony Belpaeme. The camera is capable to track moving images. }
\label{f:plate7}
\end{figure}

@@

The goal of the Talking Heads experiment is not to demonstrate 
an artificial intelligence with the same capacities as 
human intelligence, but to perform
scientific experiments so as to examine aspects 
of a theory of the origins of language and meaning. 
However, as in Turing's proposal, the public should be 
the ultimate judge whether cognitive phenomena are taking
place, and the artificial agents should be able to play their role
in interaction with humans. 
Sound scientific methodology requires that 
the experimental apparatus is a white box which can be fully probed 
by anyone who wants, that the experiments are repeatable, 
and that the phenomena that are generated 
(for example, the complexity of the lexicons or 
grammars) can be 
compared by anyone to human cognitive phenomena in order to gage 
their similarity and thus their relevance for understanding
human cognition.

@@

The {\itshape Laboratory for cognitive robots and 
teleportation} that housed the 
Talking Heads experiment also contained  
a documentation room in which the audience could 
get additional background information and 
provide feedback and commentary 
on the experiment. We created a website that was accessible 
worldwide through the Internet. This allowed viewers from anywhere
in the world to follow the dialogs, inspect the lexicons
and ontologies (sets of perceptually grounded concepts) 
of the robots, and even interact remotely with the 
physical robots, playing language games 
and doing their own experiments. We also added
other physical sites in Tokyo, Brussels, Paris, Amsterdam, 
San Jose (US) and other places, to increase the environmental complexity that 
the agents could experience.

### 2015-04-27 chap10/chap10.tex
hyphenation, punctuation, positioning, etc for chapters 1-6
hyphenations, footnotes, ...
fixed remaining typos

The main conclusion of these experiments was that the state of the art in robotics and language
game research was sufficiently advanced at that time (i.e. in 
2001) to implement these more complex games. Moreover several experiments were done to compare
different learning methods and interaction patterns, comparing social, situated learning with 
observational learning. In social situated learning, \is{social situated learning}
the learner maximally uses information from the context, the 
interaction, and the existing state of his language system, to make the best possible guess about the meaning of 
unknown utterances. He gets help from the speaker which is acting as a tutor and gives useful feedback to constrain 
the uncertainty as much as possible. An utterance is seens as purposeful and the hearer assumes that the speaker conceptualises
and expresses what is relevant to achieve communicative success. In observational learning, the learner is 
passively observing interactions of others and stores them as data. When there is a sufficiently big 
corpus he can employ cross-situational learning to grasp the meaning of each word by scanning 
through the data to find the commonalities between 
situations using the same word. This approach works also to some extend but experiments showed that it was much slower and 
not so easily made incremental \citep{DeBeule:2006}.

@@

Within this experimental setting three different strategies were tried in a long-term experiment where
a new strategy was injected at certain stages and performance monitored: 
\begin{itemize}
\item In stage \textsc{i}, agents use 
the Discrimination Game strategy and the Naming Game strategy discussed in chapters 4 and 5. A 
lexicon is emerging and there are some successful games (those for which the speaker and the hearer see the situation from 
the same perspective). But the lexicon keeps expanding and success is very limited as games where robots have a different 
perspective on the rolling ball all fail. 
\item In stage \textsc{ii}, agents have an extended 
strategy where agents can geometrically transform a model of the scene, based on their own perspective, into a model of the scene 
as seen from the perspective of the hearer. They can then conceptualise the scene from that perspective. 
Communicative success significantly increases and the 
lexicon starts to decrease. However there is ambiguity because the hearer cannot know whether
the speaker has conceptualised the 
scene from his own viewpoint or performed a perspective reversal, leading to a large number of failed communications 
(more than 50 \%). 
\item In stage \textsc{iii}, a more elaborate stratergy is used. The speaker now expresses whether he has performed perspective 
reversal (as in `{\itshape to your} left' vs. simply `left'). Only with this strategy are agents able to reach high 
communicative success with a stable inventory. 
\end{itemize}

@@

``The grammar is seen as a natural continuation of the lexicon, in the sense that it consists also of associations between forms and meanings. Use and success are monitored for each association so that the same type of self-organised coherence arises in the group, as seen in the lexicon. The form is now a more complex structure, defined as a syntactic schema. The meaning is a semantic schema. The schemas
circumscribe a feature-set in terms of a set of slots, restrictions	
on the fillers of each slot, and constraints	on the combination	
of the fillers to form the total covered by the schema. \todo{Where does this quote end?}

@@

It used the comparative method again to compare three progressively more complex strategies: 
\begin{enumerate}
\item {\bfshape  Lexical strategy}. The experiment starts from a purely 
lexical language that is predefined in all agents. The lexicon contained words for expressing properties of 
objects (such as `smooth', `red', etc.)
as well as actions involving objects (such as `move-away-from', `move-towards' or `push'). English words
were chosen to make the experiment easier to follow. Agents form
multi-word utterances using straightforward lexicon lookup, producing utterances
like `move-towards red jill' (Jill moves towards the red object) or `push 
jack jill green' (Jack pushes the green block to Jill). There is no syntax at all. Agents can nevertheless already 
achieve communicative success when they use the real world as a way to figure out the particant roles, even though the 
utterance does not provide any information. For example, to detect whether Jack pushes the green block to Jill or Jill 
pushes the green block to Jack.

\item {\bfshape  Specific marker strategy}. The second strategy introduces markers in the form of particles added to words referring 
to objects, in order to specify what agument the object fills in the predicate describing the 
event. For example, there could be two markers: `pi' 
and `pa' specifying that the referent plays the role respectively of giver and gift in a give event, as in
`pi jill pa block give'. The markers are entirely specific for each event in the sense that there is no generalisation 
yet to 'agent', 'patient' or other more abstract semantic roles. This already yields a more effective 
communication system because the complexity of seeking an interpretation matching with the world model is heavily
reduced and residual ambiguity (i.e. there is still more than one way in which the 
meaning matches with the world model) is also avoided.

@@

Here is an example of a lexical rule that defines the meaning of the word for an event `move-away-from'. The referent is 
the object that is referred to by the word, in this case it is the event itself. 
\clearpage
\begin{verbatim}
((?unit 
   (meaning
      (== (move-away-from ?event ?state)
          (move-away-from-1 ?event ?object)
          (move-away-from-2 ?event ?source)))
   (referent ?event)) 
<-->
((?unit
   (form (== move-away-from)))))
\end{verbatim}
Next, here is an example of a rule for the marker `po'. On the semantic side it ensures that the referent of one unit 
whose referent is ?source is equal to the object filling the argument move-away-from-2 in a move-away-from event. 
On the syntactic side, it triggers when there is a word that has the grammatical category `takes-marker-1' and 
is preceded by the marker `po'. (<< stands for `preceeds') 
\begin{verbatim}
((?predicate 
   (meaning
    (== (move-away-from ?evnt)
        (move-away-from-1 ?evnt ?patient)
        (move-away-from-2 ?evnt ?source))))
 (?argument
   (referent ?source))) 
<---> 
((?argument
   (gramcat (== takes-marker-1)))
 (?marker-unit
   (form (== po))) 
(<< ?marker-unit ?argument))))
\end{verbatim}

@@

The marker `po' in the above example is entirely specific for one argument of the move-away-from predicate, as would be the 
case when the second strategy is enacted. To implement the third strategy 
(to achieve more abstract semantic roles) requires that the markers become generic, which implies that 
we get additional rules that recategorise the arguments of a predicate in terms of 
semantic roles like: agent, patient, source, etc. Here is an example of a rule that 
has this effect. role-1 is an example of a semantic role created by the agent as an abstraction of the 
move-away-from-2 argument. 
\begin{verbatim}
((?predicate
   (meaning
    (== (move-away-from ?evnt) 
        (move-away-from-1 ?evnt ?patient)
        (move-away-from-2 ?evnt ?source)))))
<---> 
 ((?predicate
    (meaning
      (== (move-away-from ?evnt)
          (move-away-from-1 ?evnt ?patient)
          (move-away-from-2 ?evnt ?source)))) 
    (expanded-meaning (== (role-1 ?evnt ?source))))
\end{verbatim}
And the rules that click eveything together now operate over semantic roles as shown in the following example: 
\begin{verbatim}
 ((?predicate
   (expanded-meaning (== (role-1 ?event ?source))))
  (?unit (referent ?source))) 
<---> 
 ((?unit (gramcat (== takes-marker-1)))
  (?marker-unit (form (== po))) 
 (<< ?marker-unit ?unit))
\end{verbatim}
Recall that all rules must be reversible and that the conditional part must match and the concluding part must unify.

>>>>

The main conclusion of these experiments was that the state of the art in robotics and language
game research was sufficiently advanced at that time (i.e. in 
2001) to implement these more complex games. Moreover several experiments were done to compare
different learning methods and interaction patterns, comparing social, situated learning with 
observational learning. In social situated learning, \is{social situated learning}
the learner maximally uses information from the context, the 
interaction, and the existing state of his language system, to make the best possible guess about the meaning of 
unknown utterances. He gets help from the speaker which is acting as a tutor and gives useful feedback to constrain 
the uncertainty as much as possible. An utterance is seen as purposeful and the hearer assumes that the speaker conceptualises
and expresses what is relevant to achieve communicative success. In observational learning, the learner is 
passively observing interactions of others and stores them as data. When there is a sufficiently big 
corpus he can employ cross-situational learning to grasp the meaning of each word by scanning 
through the data to find the commonalities between 
situations using the same word. This approach works also to some extend but experiments showed that it was much slower and 
not so easily made incremental \citep{DeBeule:2006}.

@@

Within this experimental setting three different strategies were tried in a long-term experiment where
a new strategy was injected at certain stages and performance monitored: 
\begin{itemize}
\item In stage \textsc{i}, agents use 
the Discrimination Game strategy and the Naming Game strategy discussed in chapters 4 and 5. A 
lexicon is emerging and there are some successful games (those for which the speaker and the hearer see the situation from 
the same perspective). But the lexicon keeps expanding and success is very limited as games where robots have a different 
perspective on the rolling ball all fail. 
\item In stage \textsc{ii}, agents have an extended 
strategy where agents can geometrically transform a model of the scene, based on their own perspective, into a model of the scene 
as seen from the perspective of the hearer. They can then conceptualise the scene from that perspective. 
Communicative success significantly increases and the 
lexicon starts to decrease. However there is ambiguity because the hearer cannot know whether
the speaker has conceptualised the 
scene from his own viewpoint or performed a perspective reversal, leading to a large number of failed communications 
(more than 50 \%). 
\item In stage \textsc{iii}, a more elaborate strategy is used. The speaker now expresses whether he has performed perspective 
reversal (as in `{\itshape to your} left' vs. simply `left'). Only with this strategy are agents able to reach high 
communicative success with a stable inventory. 
\end{itemize}

@@

``The grammar is seen as a natural continuation of the lexicon, in the sense that it consists also of associations between forms and meanings. Use and success are monitored for each association so that the same type of self-organised coherence arises in the group, as seen in the lexicon. The form is now a more complex structure, defined as a syntactic schema. The meaning is a semantic schema. The schemas
circumscribe a feature-set in terms of a set of slots, restrictions	
on the fillers of each slot, and constraints on the combination	
of the fillers to form the total covered by the schema. \todo{This quote does not have an end}

@@

It used the comparative method again to compare three progressively more complex strategies: 
\begin{enumerate}
\item {\bfshape  Lexical strategy}. The experiment starts from a purely 
lexical language that is predefined in all agents. The lexicon contained words for expressing properties of 
objects (such as `smooth', `red', etc.)
as well as actions involving objects (such as `move-away-from', `move-towards' or `push'). English words
were chosen to make the experiment easier to follow. Agents form
multi-word utterances using straightforward lexicon lookup, producing utterances
like `move-towards red jill' (Jill moves towards the red object) or `push 
jack jill green' (Jack pushes the green block to Jill). There is no syntax at all. Agents can nevertheless already 
achieve communicative success when they use the real world as a way to figure out the participant roles, even though the 
utterance does not provide any information. For example, to detect whether Jack pushes the green block to Jill or Jill 
pushes the green block to Jack.

\item {\bfshape  Specific marker strategy}. The second strategy introduces markers in the form of particles added to words referring 
to objects, in order to specify what argument the object fills in the predicate describing the 
event. For example, there could be two markers: `pi' 
and `pa' specifying that the referent plays the role respectively of giver and gift in a give event, as in
`pi jill pa block give'. The markers are entirely specific for each event in the sense that there is no generalisation 
yet to 'agent', 'patient' or other more abstract semantic roles. This already yields a more effective 
communication system because the complexity of seeking an interpretation matching with the world model is heavily
reduced and residual ambiguity (i.e. there is still more than one way in which the 
meaning matches with the world model) is also avoided.

@@

Here is an example of a lexical rule that defines the meaning of the word for an event `move-away-from'. The referent is 
the object that is referred to by the word, in this case it is the event itself. 
\clearpage
\begin{verbatim}
((?unit 
   (meaning
      (== (move-away-from ?event ?state)
          (move-away-from-1 ?event ?object)
          (move-away-from-2 ?event ?source)))
   (referent ?event)) 
<-->
((?unit
   (form (== move-away-from)))))
\end{verbatim}
Next, here is an example of a rule for the marker `po'. On the semantic side it ensures that the referent of one unit 
whose referent is ?source is equal to the object filling the argument move-away-from-2 in a move-away-from event. 
On the syntactic side, it triggers when there is a word that has the grammatical category `takes-marker-1' and 
is preceded by the marker `po'. (<< stands for `precedes') 
\begin{verbatim}
((?predicate 
   (meaning
    (== (move-away-from ?evnt)
        (move-away-from-1 ?evnt ?patient)
        (move-away-from-2 ?evnt ?source))))
 (?argument
   (referent ?source))) 
<---> 
((?argument
   (gramcat (== takes-marker-1)))
 (?marker-unit
   (form (== po))) 
(<< ?marker-unit ?argument))))
\end{verbatim}

@@

The marker `po' in the above example is entirely specific for one argument of the move-away-from predicate, as would be the 
case when the second strategy is enacted. To implement the third strategy 
(to achieve more abstract semantic roles) requires that the markers become generic, which implies that 
we get additional rules that recategorise the arguments of a predicate in terms of 
semantic roles like: agent, patient, source, etc. Here is an example of a rule that 
has this effect. role-1 is an example of a semantic role created by the agent as an abstraction of the 
move-away-from-2 argument. 
\begin{verbatim}
((?predicate
   (meaning
    (== (move-away-from ?evnt) 
        (move-away-from-1 ?evnt ?patient)
        (move-away-from-2 ?evnt ?source)))))
<---> 
 ((?predicate
    (meaning
      (== (move-away-from ?evnt)
          (move-away-from-1 ?evnt ?patient)
          (move-away-from-2 ?evnt ?source)))) 
    (expanded-meaning (== (role-1 ?evnt ?source))))
\end{verbatim}
And the rules that click everything together now operate over semantic roles as shown in the following example: 
\begin{verbatim}
 ((?predicate
   (expanded-meaning (== (role-1 ?event ?source))))
  (?unit (referent ?source))) 
<---> 
 ((?unit (gramcat (== takes-marker-1)))
  (?marker-unit (form (== po))) 
 (<< ?marker-unit ?unit))
\end{verbatim}
Recall that all rules must be reversible and that the conditional part must match and the concluding part must unify.

### 2015-04-27 chap11/chap11.tex
fixed remaining typos

The best matching prototype for a given sensory experience is \is{prototype} 
found by a weighted sum of the match for each dimension. Many neural network models exist that perform this kind
of computation, such as Radial Basis Function networks, and their behaviour is well
understood. But the difficulty here is that robots do not have access
to a clear data set of examples and counter examples from which they
can learn how a prototype should be defined. Moreover clusters found
through unsupervised learning (such as through Kohonen
nets) may not necessarily correspond to the prototypical
views of an object. In addition, each agent independently and autonomously
develops his own repertoire of prototypes based on his history of
interactions with the world and others. Hence it is totally unlikely
that the agents have exactly the same set of prototypes, which
further aggrevates the coordination and communication problem.

@@

Given this notion of a network, the key question for designing an effective 
strategy for playing the language game becomes: What are the operators that are building and
changing the semiotic networks in each agent? We focus first on the
vocabulary, i.e. the links between nodes for individual objects $m_i$ and their 
names $f_i$. We use a variant of the strategy used in the Talking Heads experiment. 
Only the hearer changes the score after a game. 
In the case of a successful game, the score $\sigma_{f_i,m_i}$ of the used association is 
increased and its competitors are decreased according to the following
equations with the alignment rate $\gamma = 0.2$: 
\begin{equation}
 \sigma_{m_i,f_i} ≤ftarrow \sigma_{m_i,f_i} (1-\gamma) + \gamma
\end{equation}
\begin{equation}
 \sigma_{m_j,f_i} ≤ftarrow \sigma_{m_j,f_i} (1-\gamma) ~~j \neq i
\end{equation}
\begin{equation}
 \sigma_{m_i,f_k} ≤ftarrow \sigma_{m_i,f_k} (1-\gamma) ~~k \neq i
\end{equation}
A competitor is another individual $m_j$ stored in the hearer's memory 
for the name $f_i$ or another name $f_k$ for the individual 
$m_i$. When the speaker is using a name which is not linked in the network 
of the hearer to any individual, then the hearer adds a new relation between 
the recognised individual (after pointing by the speaker) and this new name. 
This alternative then competes from 
now on with existing associations through the lateral inhibition dynamics. 
We already know from many other experiments that this strategy, when used collectively in 
consecutive games between randomly chosen members of the population, leads to the  
self-organisation of a shared lexicon due to positive feedback.\footnote{
The semiotic dynamics of lateral inhibition has been studied thoroughly by Bart de Vylder as reported in his Ph.D thesis 
(2007 at the VUB AI Lab) and summarised in this paper: \cite{deVylder:2006}.}

@@

\begin{figure}[htbp]
  \centerline{\includegraphics[width=0.3\textwidth]{chap11/figs/syntactic-basic.pdf} $~~~~$ 
\includegraphics[width=0.6\textwidth]{chap11/figs/syntactic-graded.pdf}}
\caption{\label{fig:syntactic-graded} 
{\itshape Left:} Example of structure for the Basic Colour strategy. It involves three levels of hierarchy: for the word 
itself, the function of the word, and the combination of the different functions into a Colour-Phrase. Each box 
shows the name of the unit, and some of the key features of the syntactic and semantic pole.
{\itshape Right:} Example of structure for the graded colour strategy.}
\end{figure}

>>>>

The best matching prototype for a given sensory experience is \is{prototype} 
found by a weighted sum of the match for each dimension. Many neural network models exist that perform this kind
of computation, such as Radial Basis Function networks, and their behaviour is well
understood. But the difficulty here is that robots do not have access
to a clear data set of examples and counter examples from which they
can learn how a prototype should be defined. Moreover clusters found
through unsupervised learning (such as through Kohonen
nets) may not necessarily correspond to the prototypical
views of an object. In addition, each agent independently and autonomously
develops his own repertoire of prototypes based on his history of
interactions with the world and others. Hence it is totally unlikely
that the agents have exactly the same set of prototypes, which
further aggravates the coordination and communication problem.

@@

Given this notion of a network, the key question for designing an effective 
strategy for playing the language game becomes: What are the operators that are building and
changing the semiotic networks in each agent? We focus first on the
vocabulary, i.e. the links between nodes for individual objects $m_i$ and their 
names $f_i$. We use a variant of the strategy used in the Talking Heads experiment. 
Only the hearer changes the score after a game. 
In the case of a successful game, the score $\sigma_{f_i,m_i}$ of the used association is 
increased and its competitors are decreased according to the following
equations with the alignment rate $\gamma = 0.2$: 
\begin{equation}
 \sigma_{m_i,f_i} ≤\leftarrow \sigma_{m_i,f_i} (1-\gamma) + \gamma
\end{equation}
\begin{equation}
 \sigma_{m_j,f_i} ≤\leftarrow \sigma_{m_j,f_i} (1-\gamma) j \neq i
\end{equation}
\begin{equation}
 \sigma_{m_i,f_k} ≤\leftarrow \sigma_{m_i,f_k} (1-\gamma) k \neq i
\end{equation}
A competitor is another individual $m_j$ stored in the hearer's memory 
for the name $f_i$ or another name $f_k$ for the individual 
$m_i$. When the speaker is using a name which is not linked in the network 
of the hearer to any individual, then the hearer adds a new relation between 
the recognised individual (after pointing by the speaker) and this new name. 
This alternative then competes from 
now on with existing associations through the lateral inhibition dynamics. 
We already know from many other experiments that this strategy, when used collectively in 
consecutive games between randomly chosen members of the population, leads to the  
self-organisation of a shared lexicon due to positive feedback.\footnote{
The semiotic dynamics of lateral inhibition has been studied thoroughly by Bart de Vylder as reported in his Ph.D thesis 
(2007 at the VUB AI Lab) and summarised in this paper: \cite{deVylder:2006}.}

@@

\begin{figure}[htbp]
  \centerline{\includegraphics[width=0.3\textwidth]{chap11/figs/syntactic-basic.pdf} ~~~~ 
\includegraphics[width=0.6\textwidth]{chap11/figs/syntactic-graded.pdf}}
\caption{\label{fig:syntactic-graded} 
{\itshape Left:} Example of structure for the Basic Colour strategy. It involves three levels of hierarchy: for the word 
itself, the function of the word, and the combination of the different functions into a Colour-Phrase. Each box 
shows the name of the unit, and some of the key features of the syntactic and semantic pole.
{\itshape Right:} Example of structure for the graded colour strategy.}
\end{figure}
